Hydra battles in Coq

### 2020-10-27 doc/hydras.tex
correction of a fake definition of maximal-alpha-critical ordinals

Hydra games (a.k.a. \emph{Hydra battles}) appeared in an article published in 1982 by two mathematicians:
L. Kirby and J. Paris~\cite{KP82}: \emph{Accessible Independence Results for Peano Arithmetic}. 
Although the mathematical contents of this 
paper are quite advanced, the rules of hydra battles are very easy to understand. There are now several sites on Internet where you can find tutorials on hydra games, together with simulators you can play with. See, for instance, the page written by Andrej Bauer~\cite{bauer2008}.

@@

@@

Lemma lt_double : forall n:nat, n < double  n.
Proof.
   unfold double; omega.
Qed.

@@

  Definition double (n : nat) := 2 * n.

\begin{Coqsrc}
Lemma alt_double_ok n : Nat.double n = Alt.double n.
Proof.
  unfold Alt.double, Nat.double.
  omega.
Qed.
\end{Coqsrc}

@@

%\index{To do}

\begin{todo}
Look for an existing library on trees with nodes of arbitrary arity, in order to replace  this ad-hoc type with something more generic.
\end{todo}

@@

\begin{project}[**]

@@

We provide several notations for ``patterns'' which occur often in our developments.

@@

When defining a recursive function over the type \texttt{Hydra}, one has to consider the three constructors 
\texttt{node}, \texttt{hnil} and \texttt{hcons} of the mutually inductive types \texttt{Hydra} and \texttt{Hydrae}. 
Let us define for instance the function that  computes the number of nodes of any hydra:

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html}{Hydra.Hydra\_Definitions}}

@@

One may try to use the default tactic of proof by induction, that corresponds to an application of the automatically  generated  induction principle for  type \texttt{Hydra}:

@@

@@

\index{Coq!Techniques!Mutual induction}
\index{Coq!Commands!Scheme}

@@

In this section, we represent the rules of hydra battles as a binary relation associated with
a \emph{round}, i.e., an interaction composed of the two following actions:
\begin{enumerate}
\item Hercules chops off one head of the hydra
\item Then, the  hydra replicates the wounded part (if the head is at distance $\geq 2$ from the foot).
\end{enumerate}
The relation associated with each round of the battle is \emph{parameterized}  by the \emph{expected} replication  factor (irrelevant if the chopped head is at distance 1 from the foot,
but present for consistency's sake).

@@

\subsection{Chopping of a head at distance \texorpdfstring{$\geq 2$}{>= 2} from the foot (relation R2) }

@@

 Let $s$ be a sequence of hydras. 
The proposition (\texttt{S1 n s s'}) holds if $s'$ is obtained by replacing some element $h$ of $s$ by 
$n+1$ copies of $h'$, where  the proposition (\texttt{R1 h h'}) holds, in other words, \texttt{h'} is just \texttt{h}, without the chopped-off  head. \texttt{S1} is an inductive relation with two constructors that allow us to choose the position in $s'$ of the wounded sub-hydra $h$.

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html\#S1}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html}{Hydra.Hydra\_Definitions}}
\index{Coq!Type classes}
\label{types:Battle}

@@

\index{Coq!Type classes}
\label{sect:hvariant-def}

@@

\index{Coq!Type classes}

@@

\index{Coq!Type classes}
\label{types:ON}

@@

\index{Coq!Type classes}
\label{types:SubON}

@@

\index{Coq!Type classes}
\label{types:ON-for}

@@

\index{Coq!Type classes}
\label{types:ON-iso} 
\begin{Coqsrc}
Class  ON_Iso 
       `(OA : @ON A ltA compareA)
       `(OB : @ON B ltB  compareB)
       (f : A -> B)
       (g : B -> A):=
  {
  iso_compare: forall x y : A, 
      compareB (f x) (f y) = compareA x y;
  iso_inv1 : forall a, g (f a)= a;
  iso_inv2 : forall b, f (g b) = b}.
\end{Coqsrc}

@@

\subsubsection{A first attempt}
\label{sec:orgheadline77}
\index{Coq!Techniques!Well founded relations}

@@

\index{Coq!Techniques!Simultaneous induction}
\index{Coq!Techniques!Transfinite induction}

@@

Section V, 13 of~\cite{schutte} defines \emph{strongly critical} and
\emph{maximal $\alpha$-critical} ordinals.

We give their definition in \coq's syntax.

@@

Definition maximal_critical alpha : Ensemble Ord :=
  fun gamma =>
    In (Cr alpha) gamma /\
    forall ksi, alpha < ksi -> ~ In (Cr alpha) ksi.

@@

First, we define a type that contains only terms in Veblen normal form, and redefine \texttt{lt} and \texttt{compare} by delegation.

@@

Then we, prove that \texttt{lt} is a well-founded strict order and that the
function \texttt{compare} is correct.

@@

\begin{table}[h]
  \centering
  \begin{threeparttable}
    \caption{Main types and type classes}
\begin{tabular}{|c|c|c|l|}
\hline
Name & Description& Page \\\hline
Battle & Type class of battle types & \pageref{types:Battle} \\
E0 & (Well formed) ordinals below $\epsilon_0$ & \pageref{types:E0}\\
Hvariant & variants for proving termination of hydra battles & \pageref{sect:hvariant-def} \\
Hydra & tree-like representation of hydras & \pageref{types:Hydra} \\
Hydrae & finite sequences of hydras & \pageref{types:Hydrae} \\
ON & ordinal  notations & \pageref{types:ON} \\
ON\_for & comparison of an ordinal notation with Schütte's model &  \pageref{types:ON-for}\\
ON\_iso & isomorphism of ordinal notations  & \pageref{types:ON-iso}\\

>>>>

Hydra games (a.k.a. \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians:
L. Kirby and J. Paris~\cite{KP82}: \emph{Accessible Independence Results for Peano Arithmetic}. 
Although the mathematical contents of this 
paper are quite advanced, the rules of hydra battles are very easy to understand. There are now several sites on Internet where you can find tutorials on hydra games, together with simulators you can play with. See, for instance, the page written by Andrej Bauer~\cite{bauer2008}.

@@

\begin{Coqbad}
Require Import Lia.

Module Bad.

@@

Lemma lt_double : forall n:nat, n < double  n.
Proof.
   unfold double; lia.
Qed.

@@

\begin{Coqalt}
Require Import Arith Lia Iterates.
Module Alt.
  Definition double (n : nat) := iterate S n n.
End Alt.
\end{Coqalt}

\begin{Coqsrc}
Lemma alt_double_ok n : Nat.double n = Alt.double n.
Proof.
  unfold Alt.double, Nat.double; induction n; cbn.
  -  trivial.
  -  rewrite <- iterate_rw, iterate_S_eqn, <- IHn; lia.
Qed.
\end{Coqsrc}

@@

%\index{To do}
\index{Projects}

\begin{project}
Look for an existing library on trees with nodes of arbitrary arity, in order to replace  this ad-hoc type with something more generic.
\end{project}

@@

\begin{project}

@@

We provide several notations for hydra patterns  which occur often in our developments.

@@

When defining a recursive function over the type \texttt{Hydra}, one has to consider the three constructors 
\texttt{node}, \texttt{hnil} and \texttt{hcons} of the mutually inductive types \texttt{Hydra} and \texttt{Hydrae}. 
Let us define for instance the function which  computes the number of nodes of any hydra:

\vspace{4pt}
\noindent
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html}{Hydra.Hydra\_Definitions}}

@@

One may try to use the default tactic of proof by induction, which corresponds to an application of the automatically  generated  induction principle for  type \texttt{Hydra}:

@@

\vspace{4pt}
\noindent
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}}

@@

\index{Coq!Techniques!Mutually inductive types}
\index{Coq!Commands!Scheme}

@@

In this section, we represent the rules of hydra battles as a binary relation associated with
a \emph{round}, i.e., an interaction composed of the two following actions:
\begin{enumerate}
\item Hercules chops off one head of the hydra
\item Then, the  hydra replicates the wounded part (if the head is at distance $\geq 2$ from the foot).
\end{enumerate}
The relation associated with each round of the battle is parameterized  by the \emph{expected} replication  factor (irrelevant if the chopped head is at distance 1 from the foot,
but present for consistency's sake).

@@

\subsection{Chopping off a head at distance \texorpdfstring{$\geq 2$}{>= 2} from the foot (relation R2) }

@@

 Let $s$ be a sequence of hydras. 
The proposition (\texttt{S1 n s s'}) holds if $s'$ is obtained by replacing some element $h$ of $s$ by 
$n+1$ copies of $h'$, where  the proposition (\texttt{R1 h h'}) holds, in other words, $h'$ is just $h$, without the chopped-off  head. \texttt{S1} is an inductive relation with two constructors that allow us to choose the position in $s'$ of the wounded sub-hydra $h$.

\vspace{4pt}
\noindent
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html\#S1}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html}{Hydra.Hydra\_Definitions}}
\label{types:Battle}

@@

\label{sect:hvariant-def}

@@

@@

\label{types:ON}

@@

\label{types:SubON}

@@

\label{types:ON-for}

@@

\label{types:ON-iso} 
\begin{Coqsrc}
Class  ON_Iso 
       `(OA : @ON A ltA compareA)
       `(OB : @ON B ltB  compareB)
       (f : A -> B)
       (g : B -> A):=
  {
  iso_compare: forall x y : A, 
      compareB (f x) (f y) = compareA x y;
  iso_inv1 : forall a, g (f a)= a;
  iso_inv2 : forall b, f (g b) = b}.
\end{Coqsrc}

@@

\subsubsection{A first attempt}
\label{sec:orgheadline77}
\index{Coq!Techniques!Well-founded induction}

@@

\index{Coq!Techniques!Transfinite induction}

@@

Section V, 13 of~\cite{schutte} defines \emph{strongly critical} and
\emph{maximal $\alpha$-critical} ordinals:

\begin{itemize}
\item $\alpha$ is strongly critical if
$\alpha$ is $\alpha$-critical,
\item $\gamma$ is maximal $\alpha$-critical if $\gamma$ is $\alpha$-critical, and, for all $\xi>\alpha$, $\gamma$ is not $\xi$-critical.

@@

Definition maximal_critical alpha : Ensemble Ord :=
  fun gamma =>
    In (Cr alpha) gamma /\
    forall xi, alpha < xi -> ~ In (Cr xi) gamma.

@@

First, we define a type that contains only terms in Veblen normal form, and redefine \texttt{lt} and \texttt{compare} by delegation (see for comparison the construction of type \texttt{E0} in Sect.~\vref{sect:E0-def}).

@@

Then, we prove that \texttt{lt} is a well-founded strict order and that the
function \texttt{compare} is correct.

@@

\begin{table}[h]
  \centering
  \begin{threeparttable}
    \caption{Main types and type classes}
\begin{tabular}{|c|c|c|l|}
\hline
Name & Description& Page \\\hline
Battle & type class of battle types & \pageref{types:Battle} \\
E0 & (well formed) ordinals below $\epsilon_0$ & \pageref{types:E0}\\
Hvariant & variants for proving termination of hydra battles & \pageref{sect:hvariant-def} \\
Hydra & tree-like representation of hydras & \pageref{types:Hydra} \\
Hydrae & finite sequences of hydras & \pageref{types:Hydrae} \\
ON & ordinal  notations & \pageref{types:ON} \\
ON\_for & comparison of an ordinal notation with Schütte's model &  \pageref{types:ON-for}\\
ON\_iso & isomorphism of ordinal notations  & \pageref{types:ON-iso}\\
